Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-05-17 analysis/paper/abstract.Rmd
revise project structure to comply with dissertation

# Abstract

<!-- One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline -->

<!-- Two to three sentences of more detailed background, comprehensible to scientists in related disciplines -->

<!-- One sentence clearly stating the general problem being addressed by this particular study -->

<!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->

<!-- Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge -->

<!-- One or two sentences to put the results into a more general context -->

<!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion. Under these circumstances, the length of the paragraph can be yp to 300 words. (This example is 190 words without the final section, and 250 words with it).

>>>>

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.
